Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data

### 2021-01-18 paper/paper.Rmd
New subsection, more discussion, reply AE

To understand why points tend to be away from the center in the high-dimensional space, but crowd the center in low-dimensional projections, it is helpful to consider the projected volume relative to high-dimensional volume. To avoid edge effects and to impose rotation invariance, we will start from the data being uniformly distributed in a hypersphere, i.e. all data points are within a specified distance from the center. This makes calculations more tractable than assuming a uniform distribution in hypercube (box).

@@

@@

An alternative is the slice tour [@sliceTour] which  allows distributions
of points around the center of the data to be explored using sections instead of projections. The slice tour is useful when there are large numbers of observations
or if there is concave structure in the data.

>>>>

To understand why points tend to be away from the center in the high-dimensional space, but crowd the center in low-dimensional projections, it is helpful to consider the projected volume relative to high-dimensional volume. To avoid edge effects and to impose rotation invariance, we start from the volume in a hypersphere, i.e. the volume is contained within a specified distance from the center. This is more tractable than assuming a hypercube (box).

<!-- To avoid edge effects and to impose rotation invariance, we will start from the data being uniformly distributed in a hypersphere, i.e. all data points are within a specified distance from the center. This makes calculations more tractable than assuming a uniform distribution in hypercube (box).-->

@@

## Data preprocessing {#sec:prep}

Before applying the transformation in practice, it can be important to preprocess the data, as generally required with methods for the visualisation of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation.

Scaling is commonly also combined with centering as a preprocessing step. This is less important here where we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane).

Even relatively low numbers of dimensions, see e.g. examples with $p=10$, already lead to sizable piling and thus large distortions in the sage display. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it makes sense to first reduce the number of dimensions, for example using principal components analysis, and to adjust the $\gamma$ parameter if needed.

We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of data distributions. XXX specific examples? explicitly mention other transformations?

@@

An alternative is the slice tour [@sliceTour] which  allows distributions
of points around the center of the data to be explored using sections instead of projections.
\textcolor{red}{Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.}
The slice tour is useful when there are large numbers of observations or if there is concave structure in the data.
\textcolor{red}{Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation. In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.}

\textcolor{red}{As seen in the examples, the new sage display is especially useful to circumvent the issue of overplotting. Several other approaches exist to address this problem, notably also the use of density displays or the use of transparancy in a scatterplot display. Note however that these will in general still suffer from data piling and thus reduced resolution near the center of the projection. These approaches could therefore be limited for some of the tasks discussed here, for example the identification of class boundaries or the visualisation of small features near hidden near the center of a distribution.}

### 2021-01-18 paper/paper.tex
New subsection, more discussion, reply AE

To understand why points tend to be away from the center in the
high-dimensional space, but crowd the center in low-dimensional
projections, it is helpful to consider the projected volume relative to
high-dimensional volume. To avoid edge effects and to impose rotation
invariance, we will start from the data being uniformly distributed in a
hypersphere, i.e.~all data points are within a specified distance from
the center. This makes calculations more tractable than assuming a
uniform distribution in hypercube (box).

@@

@@

An alternative is the slice tour \citep{sliceTour} which allows
distributions of points around the center of the data to be explored
using sections instead of projections. The slice tour is useful when
there are large numbers of observations or if there is concave structure
in the data.

>>>>

To understand why points tend to be away from the center in the
high-dimensional space, but crowd the center in low-dimensional
projections, it is helpful to consider the projected volume relative to
high-dimensional volume. To avoid edge effects and to impose rotation
invariance, we start from the volume in a hypersphere, i.e.~the volume
is contained within a specified distance from the center. This is more
tractable than assuming a hypercube (box).

@@

\hypertarget{sec:prep}{%
\subsection{Data preprocessing}\label{sec:prep}}

Before applying the transformation in practice, it can be important to
preprocess the data, as generally required with methods for the
visualisation of high-dimensional data. In particular, when working with
projections that combine multiple variables it is important to ensure a
common scale across the dimensions. This is especially important with
the sage display, where the overall scale (set by the maximum radius
\(R\)) determines the effect of the transformation.

Scaling is commonly also combined with centering as a preprocessing
step. This is less important here where we always center the projected
data before applying our transformation (i.e.~the projected mean defines
the origin of our polar coordinate system in the plane).

Even relatively low numbers of dimensions, see e.g.~examples with
\(p=10\), already lead to sizable piling and thus large distortions in
the sage display. However, we are often confronted with much higher
dimensional datasets in which many of the variables are primarily adding
noise. In this case it makes sense to first reduce the number of
dimensions, for example using principal components analysis, and to
adjust the \(\gamma\) parameter if needed.

We may also wish to apply non-linear transformations before visualising
the data or combining variables in a projection. One example would be to
use a log transformation in the case of a skewed distribution. These
types of transformations will generally change the distribution across
the high-dimensional volume (and in particular also the position of the
mean), and can thus make the sage display more useful for a wide range
of data distributions. XXX specific examples? explicitly mention other
transformations?

@@

An alternative is the slice tour \citep{sliceTour} which allows
distributions of points around the center of the data to be explored
using sections instead of projections.
\textcolor{red}{Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.}
The slice tour is useful when there are large numbers of observations or
if there is concave structure in the data.
\textcolor{red}{Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation. In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.}

\textcolor{red}{As seen in the examples, the new sage display is especially useful to circumvent the issue of overplotting. Several other approaches exist to address this problem, notably also the use of density displays or the use of transparancy in a scatterplot display. Note however that these will in general still suffer from data piling and thus reduced resolution near the center of the projection. These approaches could therefore be limited for some of the tasks discussed here, for example the identification of class boundaries or the visualisation of small features near hidden near the center of a distribution.}
